Quality Criteria for SKOS Vocabularies

### 2012-03-21 criteria.tex
fixed some todos

@@

SKOS defines a set of properties that link resources with RDF Literals, which are plain text natural language strings with an optional language tag. This includes the labeling properties \texttt{skos:prefLabel}, \texttt{skos:altLabel}, \texttt{skos:hiddenLabel} and also SKOS documentation properties, such as \texttt{skos:note} and subproperties thereof. Since omitting language tags or using non-standardized, private language tags in a SKOS vocabulary could limit language-dependent structural queries, we check a SKOS vocabulary for omitted language tags by iterating over resources in $IR$ and collect those that have relations to plain literals without or with invalid language tags. We delegate validity checking of language tags to methods provided by the by OpenRDF\footnote{\url{http://www.openrdf.org/}} framework that in turn checks for conformance with RFC3066\footnote{\url{http://www.ietf.org/rfc/rfc3066.txt}}.\todo{CM}{is there any literature reference for this criterion?}

@@

A SKOS vocabulary that uses language tags with its literal values, should apply them consistently for all concepts. Hence, the set of language tags used by the literal values linked with a \todo{CM}{concept/resource} should be the same for all \todo{CM}{concepts/resources}. This is particularly important for applications that rely on internationalization and translation use cases. We calculate the affected resources by extracting the global set of language tags used in a vocabulary from all literal values attached to all \todo{CM}{concepts/resources}. Then we iterate over all \todo{CM}{conceptual resources} in $C$ and identify all conceptual resources having a set of language tags that is not equal to the global language set. \todo{CM}{literature reference?}

@@

\cite{Aitchison2000} cites Svenonius \todo{CM}{cite original work} to advocate ``inclusion of as much definition material as possible''. The SKOS language specification defines various properties that are intended to hold this kind of information, subsumed as ``documentation properties'' in the SKOS reference. To identify the affected conceptual resources, we iterate over all elements in $C$ to find those that are not using at least one of the documentation properties.\todo{CM}{rephrase}

\paragraph{QC 4: Potentially Semantically Related Concepts} \todo{All}{reformulate criterion title: Labeling Conflicts}

% avoid "given the fact, " sentences -> rephrase

This criterion is an extension of a statement from the SKOS primer, recommending ``no two concepts have the same preferred lexical label in a given language when they belong to the same concept scheme''~[\todo{CM}{ref}]. Given the fact that (i) many vocabularies don’t make use of concept schemes and (ii) also identical \texttt{skos:altLabels} or \texttt{skos:hiddenLabels} can have a negative effect in some use-cases (e.g, auto-completion of concept labels based on user input), we follow a more general approach: identification of all pairs of concepts with their respective \texttt{skos:prefLabel}, \texttt{skos:altLabel} or \texttt{skos:hiddenLabel} meeting a threshold of a certain similarity function $sim:LV \times LV \rightarrow [0,1]$. For the analysis of the criterion in this document, we define the similarity function  to check for case-insensitive string equality and the threshold to equal 1. Calculation of this criterion is done by applying the similarity function to every label of every possible pair \todo{CM}{crossproduct?} of conceptual concepts in $C$ and filtering those that meet or exceed a specified threshold.

@@

\subsubsection{Cyclic Hierarchical Relations} have been mentioned numerous times in thesaurus development literature. \cite{Soergel2002} suggests a ``check for hierarchy cycles" since they ``throw the program for a loop in the generation of a complete hierarchical structure''. However, SKOS does not define a formal constraint regarding hierarchy cycles but the SKOS reference document mentions them as potential problems. The SKOS documentation also does not state, how exactly hierarchical relationships are to be interpreted. There exist common forms like, e.g., ``generic-specific'', ``instance-of'' or ``whole-part'' \cite{Hedden2010,Harpring2010,Aitchison2000} where cycles would be considered a logical contradiction. To calculate this criterion, we construct two graphs, with the set of nodes being $C$. The edges in one graph only point towards the broader concepts, the edges of the other graph towards the narrower concepts. For each graph we now identify those nodes that are part of a cycle and return the set of strongly connected components they are contained in.

@@

\subsubsection{Solely Transitively Related Concepts}
Two concepts are related using only transitive hierarchical relations which are, according to the SKOS reference document, ``not used to make assertions''. Transitive hierarchical relations in SKOS are meant to be infered by the vocabulary user, which is reflected in the SKOS schema by, e.g., \texttt{broader} being a subproperty of \texttt{broaderTransitive}. This criterion identifies the set $I_{EXT}(\texttt{skos:broaderTransitive}^\mathcal{I}) \cup I_{EXT}(\texttt{skos:narrowerTransitive}^\mathcal{I})$ \textbf{without} RDFS subproperty entailment.

>>>>

\todo{All}{decide for either using "concept" or "conceptual resource"}

@@

SKOS defines a set of properties that link resources with RDF Literals, which are plain text natural language strings with an optional language tag. This includes the labeling properties \texttt{skos:prefLabel}, \texttt{skos:altLabel}, \texttt{skos:hiddenLabel} and also SKOS documentation properties, such as \texttt{skos:note} and subproperties thereof. Since omitting language tags or using non-standardized, private language tags in a SKOS vocabulary could limit language-dependent structural queries, we check a SKOS vocabulary for omitted language tags by iterating over resources in $IR$ and collect those that have relations to plain literals without or with invalid language tags. We delegate validity checking of language tags to methods provided by the by the OpenRDF\footnote{\url{http://www.openrdf.org/}} framework that in turn checks for conformance with RFC3066\footnote{\url{http://www.ietf.org/rfc/rfc3066.txt}}.\todo{CM}{is there any literature reference for this criterion?}

@@

A SKOS vocabulary that uses language tags with its literal values, should apply them consistently for all concepts. Hence, the set of language tags used by the literal values linked with a concept should be the same for all concepts. This is particularly important for applications that rely on internationalization and translation use cases. We calculate the affected concepts by extracting the global set of language tags used in a vocabulary from all literal values attached to all concepts in $C$. In a second pass we again iterate over this set of concepts and identify all conceptual resources having a set of language tags that is not equal to the global language set. \todo{CM}{literature reference?}

@@

\cite{Svenonius1997} advocates the ``inclusion of as much definition material as possible''. The SKOS language specification defines various properties that are intended to hold this kind of information, subsumed as ``documentation properties''~\cite{SkosReference2008}. To identify all undocumented concepts, we iterate over all concepts in $C$ and collect those for which we cannot find information linked by any of the documentation properties.

\paragraph{QC 4: Label Conflicts}

This criterion is an extension of a statement from \cite{Isaac2009}, recommending ``no two concepts have the same preferred lexical label in a given language when they belong to the same concept scheme''. Based on our review of existing vocabularies, we could observe that (i) many vocabularies don’t make use of concept schemes and (ii) also identical \texttt{skos:altLabels} or \texttt{skos:hiddenLabels} can have a negative effect in some use-cases (e.g, auto-completion of concept labels based on user input). Thus we follow the more general approach of finding all pairs of concepts with their respective \texttt{skos:prefLabel}, \texttt{skos:altLabel} or \texttt{skos:hiddenLabel} meeting a threshold of a certain similarity function $sim:LV \times LV \rightarrow [0,1]$. For the analysis of the criterion in this document, we define the similarity function to check for case-insensitive string equality and the threshold to equal 1. Calculation of this criterion is done by iterating over all pairs of concepts, i.e., elements of $C \times C$. We then apply the $sim$ to every possible combination of labels between the members of the pairs and collect those pairs for which $sim$ meets or exceeds a specified threshold.

@@

\subsubsection{Cyclic Hierarchical Relations} have been mentioned numerous times in thesaurus development literature. \cite{Soergel2002} suggests a ``check for hierarchy cycles" since they ``throw the program for a loop in the generation of a complete hierarchical structure''. However, SKOS does not define a formal constraint regarding hierarchy cycles but \cite{SkosReference2008} mentions them as potential problems. The SKOS documentation also does not state, how exactly hierarchical relationships are to be interpreted. There exist common forms like, e.g., ``generic-specific'', ``instance-of'' or ``whole-part'' \cite{Hedden2010,Harpring2010,Aitchison2000} where cycles would be considered a logical contradiction. To calculate this criterion, we construct two graphs, with the set of nodes being $C$. The edges in one graph only point towards the broader concepts, the edges of the other graph towards the narrower concepts. For each graph we now identify those nodes that are part of a cycle and return the set of strongly connected components they are contained in.

@@

\subsubsection{Solely Transitively Related Concepts}
Two concepts are related using only transitive hierarchical relations which are, according to \cite{SkosReference2008}, ``not used to make assertions''. Transitive hierarchical relations in SKOS are meant to be infered by the vocabulary user, which is reflected in the SKOS schema by, e.g., \texttt{broader} being a subproperty of \texttt{broaderTransitive}. This criterion identifies the set $I_{EXT}(\texttt{skos:broaderTransitive}^\mathcal{I}) \cup I_{EXT}(\texttt{skos:narrowerTransitive}^\mathcal{I})$ \textbf{without} RDFS subproperty entailment.
